Machine Learning guided Design\\ of Ribosome Binding Sites

### 2021-06-15 paper/main.tex
Updates from Overleaf

% Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
Ribosome binding site (RBS), as one of the engineering transcription and translation control elements, allows fine control of gene expression.
% Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
We consider the problem of designing RBSs in nucleotides level to optimise the gene expression with fixed budget. 
To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
We have integrated these ML algorithms with laboratory automation and high-throughput processes for more reliable data generation.
Our approach has shown that the algorithms were able to identify rules conducive to high protein expression.
Notably, by testing only 450 RBS variants, we have discovered 
% designed and experimentally-validated 
RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Overall, our results show that machine learning is a very powerful tool in design of RBSs and pave the way for the experimental design of more complicated genetic devices.

@@

The design recommendations were made using the Multi-armed Bandit algorithm.
In short, this algorithm is a stochastic method of probing of the experimental space. 
It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
It also balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\

@@

In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).\\

@@

Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round. 
As expected, the predictions in Round 0 were poor due to use of approximated data. 
The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.\\

One important point to note is that the predictions are also influenced by our recommendation choices. 
In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having real mean different to what was predicted.
However, this is still very useful information for future predictions as it allows us to understand the underlying space better.\\

@@

\begin{table}[!h]
\centering
\begin{minipage}[c]{0.6\textwidth}
\centering
\begin{tabular}{|c|c|}
\hline
\textbf{Characteristics of the library}                                                       & \textbf{Statistics} \\ \hline
Total experimental space                                                                      & 4096                \\ \hline
Planned constructs                                                                            & 450                 \\ \hline
Successfully constructed                                                                      & 445                 \\ \hline
Sequences with CV\textless{}40\%                                                              & 79\%                \\ \hline
Sequences with CV\textless{}20\%                                                              & 27\%                \\ \hline
\begin{tabular}[c]{@{}c@{}}Efficiency of bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
Raw TIR range                                                                                     &      [4.926, 105.377]               \\ \hline
TIR ratio range                                                                                     &      [0.063, 1.339]               \\ \hline
\end{tabular}
\end{minipage}
\begin{minipage}[c]{0.38\textwidth}
\centering
% Please add the following required packages to your document preamble:
% \usepackage[table,xcdraw]{xcolor}
% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
\begin{tabular}{|c|c|}
\hline
\textbf{Top RBS Core} & \textbf{TIR Ratio} \\ \hline
GGGGGC                & 1.339             \\ \hline
GGGGGT                & 1.154             \\ \hline
GGCTAT                & 1.084              \\ \hline
\textbf{AGGAGA}                & 1                  \\ \hline
GGCGTT                & 0.981            \\ \hline
GGGGGG                & 0.979             \\ \hline
GGCGAC                & 0.976             \\ \hline
CAGGAG                & 0.963             \\ \hline
GGCGAG                & 0.952             \\ \hline
\textbf{AGGAGG}                & 0.394            \\ \hline
\end{tabular}
\end{minipage}
\caption{\textbf{Characteristics of the library.}
Left table presents some of the characteristics of our library.
Right table presents 10 RBS sequences with their corresponding TIR ratios, the first 9 being the strongest sequences including the benchmark sequence (AGGAGA) and the last being the Shine-Dalgarno sequence (AGGAGG).
CV is coefficient of variation (STD of a sample divided by its mean).
Efficiency of the bandit design is calculated by dividing the highest TIR found using machine learning by the highest TIR found using random sequence generation. }
\end{table}

Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high. 
We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
Figure \ref{fig:Library characteristics}B shows what edit distance is required for positive change in TIR for RBS with TIR $>0.75$.
For RBS with high TIR ($>1$), the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
For RBS with medium TIR ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
That means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
This also implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.  
In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
In such cases, a directed process should be strongly encouraged.
This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed \cite{Jackel2008}.\\

Finally, our strong sequences did not show neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS nor any obvious secondary structures that could explain their TIRs (see Supplementary).
This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \cite{Saito2020,EspahBorujeni2016} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the TIR function to be optimised and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
% We have investigated a number of methods of digitising the DNA sequence, finally settling on 
To represent RBSs and capture the similarities between them, we chose to use the
Weighted Degree Kernel with Shift method, which fit well into our prediction method.
In building and testing, we have performed bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow and testing only 450 RBS variants, we have discovered 
% designed and experimentally-validated 
RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBS that can be used in the future studies.\\

Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require a substantial time and experiments to uncover.
For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR of such sequences.
We have achieved this despite the relatively low accuracy of our predictions, which benefits from recommendation algorithms with the exploitation-exploration balance.
This gives hope that the continuing rise in use of machine learning in synthetic biology will enable to uncover new biologically relevant rules of design.\\

We have found our approach bringing machine learning and synthetic biology experts very powerful.
We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
Synthetic biology is promoting standardised and normalised testing in biology which naturally pairs with machine learning which can leverage the high quality biological data sets generated when the correct design rules are followed.
The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
On top of that, the use of automation has helped us to produce more reliable results, which gave us the required confidence in our predictions and recommendations.\\

In this study we have limited the number of design rounds to four.
There was a number of reasons for this, including time and money limitations, but also the results obtained showed that we have achieved our goal of generating very strong RBS designs.
There is a possibility that increasing number of experimental rounds would enable us to improve the results further, however this has to be put in the context of the limited resources.
For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would need unreasonable use of resources.
Our approach also compares favourably with approaches using deep learning \cite{Hollerer2020}, where the data sets required for accurate predictions are 3 orders of magnitude bigger than ours. \\

There are still open questions that need to be addressed for applying machine learning in synthetic biology.
Firstly, we would like to understand how can we extract more biologically important information from the decisions made by our algorithms.
We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the obtained results.
Secondly, given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
Thirdly, the generalisability of the method is unknown.
We believe that the method described here would be useful for use in design of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence and so the method's applicability might be impacted at some point.
Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.
In this research we have decided to gradually move the balance from exploration to exploitation over the rounds, but it is not yet clear what is the optimal way of conducting this change.\\

In the future, we hope to extend the algorithm to other, more complicated genetic elements.
This could include, for example, promoters and terminators.
However, it's important to reiterate that the complexity of the task quickly increases with the length of the sequence.
This is because the experimental space grows exponentially with the number of examined positions and so the space becomes increasingly hard to cover with experiments.
To solve this problem, a different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\

>>>>

% Fine control of gene expression can be achieved through engineering transcription and translation control elements, including the ribosome binding site (RBS).
The ribosome binding site (RBS), as one of the transcription and translation control elements, allows fine control of gene expression.
% Unfortunately, RBSs are not understood at the level of finesse required for reliable design.
We consider the problem of designing RBSs at the nucleotide level to optimise gene expression with a fixed budget. 
To address this problem, we have created a machine learning (ML) guided Design-Build-Test-Learn (DBTL) cycle for the experimental design of bacterial RBSs.
We used Gaussian Process Regression for prediction and the batch Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vivo.
We have integrated these machine learning algorithms with laboratory automation and high-throughput processes for more reliable data generation.
The algorithms were able to identify RBS sequences rules conducive to high protein expression.
Notably, by testing only 450 RBS variants, we have discovered 
% designed and experimentally-validated 
RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Overall, our results show that machine learning is a very powerful tool for designing RBSs, and they pave the way for the design of more complicated genetic devices.

@@

@@

In the subsequent 3 rounds, with one batch each, all 90 designs were generated using our machine learning algorithm based on the data obtained from the previous rounds (these groups are called Bandit 1 to 3 respectively).
The bandit recommendations were made using the batch Upper Confidence Bound algorithm.
% In short, this algorithm is a stochastic method of probing of the experimental space. 
% It aims at maximising the reward (output) from testing a limited number of instances from a large pool which cannot be tested exhaustively due to limited resources (time, computational power, money).
It balances the exploration-exploitation paradigm, where exploration focuses on testing data points which maximise information gain and exploitation focuses on recommending RBSs with high predicted TIR.\\

@@

Figure \ref{fig: Scatterplot} shows how our implementation of the Gaussian Process algorithm performed in terms of predictions in each round. 
As expected, the predictions in Round 0 were poor due to use of approximated data. 
The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.067 for round 0 to R\textsuperscript{2} of 0.213 for round 3.
Similarly, the Spearman correlation coefficient rose from 0.269 for Round 0 to 0.546 for Round 3.\\

One important point to note is that the predictions are also influenced by our recommendation choices. 
In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having a real mean different from what was predicted.
However, this is still very useful information for future predictions as it allows us to understand the underlying space better.\\

@@

\begin{table}[!h]
\centering
\begin{minipage}[c]{0.6\textwidth}
\centering
\begin{tabular}{|c|c|}
\hline
\textbf{Characteristics of the library}                                                       & \textbf{Statistics} \\ \hline
Total experimental space                                                                      & 4096                \\ \hline
Planned constructs                                                                            & 450                 \\ \hline
Successfully constructed                                                                      & 445                 \\ \hline
Sequences with CV\textless{}40\%                                                              & 79\%                \\ \hline
Sequences with CV\textless{}20\%                                                              & 27\%                \\ \hline
\begin{tabular}[c]{@{}c@{}}Efficiency of bandit design \\ (compared with random)\end{tabular} & 2                   \\ \hline
Raw TIR range                                                                                     &      [4.93, 105.38]               \\ \hline
TIR ratio range                                                                                     &      [0.06, 1.34]               \\ \hline
\end{tabular}
\end{minipage}
\begin{minipage}[c]{0.38\textwidth}
\centering
% Please add the following required packages to your document preamble:
% \usepackage[table,xcdraw]{xcolor}
% If you use beamer only pass "xcolor=table" option, i.e. \documentclass[xcolor=table]{beamer}
\begin{tabular}{|c|c|}
\hline
\textbf{Top RBS Core} & \textbf{TIR Ratio} \\ \hline
GGGGGC                & 1.34             \\ \hline
GGGGGT                & 1.15             \\ \hline
GGCTAT                & 1.08              \\ \hline
\textbf{AGGAGA}                & 1                  \\ \hline
GGCGTT                & 0.98            \\ \hline
GGGGGG                & 0.98             \\ \hline
GGCGAC                & 0.98             \\ \hline
CAGGAG                & 0.96             \\ \hline
GGCGAG                & 0.95             \\ \hline
\textbf{AGGAGG}       & 0.39            \\ \hline
\end{tabular}
\end{minipage}
\caption{\textbf{Characteristics of the library.}
Left table presents some of the characteristics of our library.
Right table presents 10 RBS sequences with their corresponding TIR ratios, the first 9 being the strongest sequences including the benchmark sequence (AGGAGA) and the last being the Shine-Dalgarno sequence (AGGAGG).
CV is coefficient of variation (STD of a sample divided by its mean).
Efficiency of the bandit design is calculated by dividing the highest TIR found using machine learning by the highest TIR found using random sequence generation. }
\end{table}

Another interesting characteristic uncovered by our research is the perceived editing distance between two sequences required for  improvement in the TIR when the given RBS' TIR is already high. 
We define the edit distance as Hamming distance, that is, how many positions have to be changed to get from one sequence to the other (Hamming distance of 0 means that the sequences are identical and 6 means that they are two completely different sequences).
Figure \ref{fig:Library characteristics}B shows what edit distance is required for positive change in TIR for RBS with TIR $>0.75$.
For RBS with high TIR ($>1$), the minimum distance that is required for increase of TIR is 2, with edit distance between 2 and 5 giving similar results.
For RBSs with medium TIRs ($<1$), a distance of 1 is enough to produce a meaningful increase in TIR.
This means that as the TIR of examined RBSs increases, exploring sequences which are more dissimilar to the current candidates tends to give more meaningful improvement. 
This also implies that the low rate of natural mutations will be very slow to explore more dissimilar sequences on such a short distance \cite{Lee2012}, which indicates that methods like Adaptive Laboratory Evolution may not be able to find very strong RBSs within limited budget.  
In other words, because the examined sequence is relatively short (6bp in a wider 20bp context) the time to accumulate 2 or more changes in the RBS region required for meaningful increase in TIR might be prohibitively long.
In such cases, a directed process should be strongly encouraged.
This is in line with common practices in e.g. protein engineering, where similar approaches, that is making more directed changes, are often observed \cite{Jackel2008}.\\

Finally, our strong sequences showed neither strong binding to the anti-sense sequence of the ribosome known to bind to RBS, nor any obvious secondary structures that could explain their TIRs (see Supplementary).
This result combined with the unexpectedly bimodal nature of KDEs in Figure \ref{fig: Swarmplot and Quantplot} reinforces the notion based on the previously reported literature \cite{Saito2020,EspahBorujeni2016} that there may be a number of different mechanisms governing the probability of effective RBS-ribosome binding.\\

@@

In this work, we have shown how machine learning and high-throughput, automated laboratory methods can be jointly applied to efficiently generate a library of small parts, in this case bacterial RBS. 
We have used Gaussian Process regression to predict the TIR function to be optimised and Upper Confidence Bound Bandit algorithm to recommend sequences to be tested.
% We have investigated a number of methods of digitising the DNA sequence, finally settling on 
To represent RBSs and capture the similarities between them, we chose to use the
Weighted Degree Kernel with Shift method, which fit well into our prediction method.
In building and testing, we have performed the bulk of our experiments using automation to increase their speed, reliability and reproducibility.
By using our workflow and testing only 450 RBS variants, we have discovered 
% designed and experimentally-validated 
RBSs with high translation initiation rates equalling or exceeding the currently known strong RBSs by up to 34\%.
Furthermore, we have generated an extensive library of diverse RBSs that can be used in future studies.\\

Our approach has shown that machine learning is able to correctly detect and exploit rules of biological design that otherwise require substantial time and experiments to uncover.
For example, our algorithm has correctly identified the correlation of high guanine content in the RBS with high TIR.
We have achieved this despite the relatively low accuracy of our predictions, which benefits from recommendation algorithms with the exploitation-exploration balance.
This gives hope that the continuing rise in use of machine learning in synthetic biology will enable us to uncover new biologically relevant rules of design.\\

We have found our approach of bringing machine learning and synthetic biology experts together very powerful.
We envision that pairing machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
Synthetic biology is promoting standardised and normalised testing in biology which naturally pairs with machine learning which can leverage the high quality biological data sets generated when the correct design rules are observed.
The addition of machine learning to synthetic biology also adds an additional layer of scrutiny to the generated data sets through the advanced statistical methods that can be used to design and analyse the experiments.
On top of that, the use of automation has helped us to produce more reliable results, which gave us the required confidence in our predictions and recommendations.\\

In this study we have limited the number of design rounds to four.
There was a number of reasons for this, including time and money limitations, but also the results obtained showed that we have achieved our goal of generating very strong RBS designs.
There is a possibility that increasing the number of experimental rounds would enable us to improve the results further, however this has to be put in the context of the limited resources.
For example, scanning the whole space would surely achieve the best results, i. e. would enable us to find the strongest possible RBS, but that would require unreasonable use of resources.
Our approach also compares favourably with approaches using deep learning \cite{Hollerer2020}, where the data sets required for accurate predictions are 3 orders of magnitude bigger than ours. \\

There are still open questions that need to be addressed for applying machine learning in synthetic biology.
Firstly, we would like to understand how we can extract more biologically-important information from the decisions made by our algorithms.
We have shown that the algorithms are able to exploit them, but it will be important to create tools that will enable their reliable extraction from the obtained results.
Secondly, given the small number of RBS sequences tested, how can machine learning algorithms provide more accurate predictions and uncertainty measurement? 
Thirdly, the generalisability of the method is unknown.
We believe that the method described here would be useful for designing other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence, so the method's applicability might be impacted at some point.
Finally, the optimal exploration-exploitation balance between rounds and samples is still to be determined.
In this research we have decided to gradually move the balance from exploration to exploitation across the rounds, but it is not yet clear what is the optimal way of conducting this change.\\

In the future, we hope to extend the algorithm to other more complicated genetic elements, including promoters and terminators.
However, it is important to reiterate that the complexity of the task quickly increases with the length of the sequence.
This is because the experimental space grows exponentially with the number of examined positions, so the space becomes increasingly hard to cover with experiments.
To solve this problem, different algorithms or experimental techniques might be needed, but the general workflow can be reused.\\
